Searching for circuit complexity lower bounds using SAT solvers

### 2022-05-14 report/draft1.tex
finished report

If functions with circuits of a given complexity could be identified by an efficient algorithm, this could be used to construct efficient learning algorithms for circuits of a smaller complexity; this would resolve one of the biggest questions in learning theory~\cite{carmosino}. Furthermore, such an algorithm could be used to break pseudorandom generators (i.e. distinguish random from pseudorandom inputs)~\cite{razborov94}. As a result, this problem has strong connections in both learning theory and cryptography.

A heuristic approach is to find minimal circuits for small instances of the functions. This can be used to prove new bounds~\cite{kulikovsurvey} and may lead to theoretical insights on the structure of their optimal circuits generally~\cite{williams}. Doing this for small functions computing natural like parity, modulo and addition allows us to gain a greater understanding of their difficulty and compute them more efficiently when using low-level code or hardware.

Finding efficient circuits is also necessary in practice when designing electronic systems. Electronic circuit design is an extensive field of its own, involving circuit simulation software to test designs, or hardware specification languages such as Verilog and circuit synthesis software to convert specifications into circuits. However, electronic circuit design is concerned with a wider range of problems (for example, functions whose output changes over time) and with concerns about correct physical implementation; we focus on abstract circuits.

@@

Circuits are a popular model of computation; they are non-uniform, meaning that in a family of circuits where each takes a different size of input, each circuit is independent and not constructed by a common algorithm. Notably, this means that circuits are not subject to the Space and Time Hierachy Theorems, although there is a Non-Uniform Hierarchy Theorem stating that larger circuits can compute strictly more functions than smaller ones~\cite{arora}. In the past there have been theoretical proofs of lower bounds for restricted circuit classes, such as $AC^0$ (functions computed by circuit families of $O(1)$ depth)~\cite{furst81}~\cite{ajtai83}, $AC^0[k]$ (i.e. $AC^0$ additionally with MOD-k gates)~\cite{razborov87}~\cite{smolensky87}, and monotone circuits~\cite{razborov85}. There was also a bound showing the practical impossibility of deciding bounded-size formulae in a certain second-order language~cite{stockmeyer02}. However, no lower bounds better than (3+1/86)n-o(n) have been found for general circuits over the basis $B_2$ consisting of all functions on 2 bits~\cite{find16}, while the best bound is 5n-c for the basis $U_2 = \{ \land, \lor, \neg\}$~\cite{lachish01}~\cite{iwama02}. Furthermore, there are results stating that it is impossible to use circuit lower bounds to separate P and NP by relativising methods~\cite{baker75}, algebrizing methods~\cite{aaronson09}, or (under standard cryptographic assumptions) so-called `natural proofs'~\cite{razborov94}. A formula in Disjunctive Normal Form (DNF) is one consisting of clauses (disjunctions of literals), with the final formula consisting of the conjunction of all clauses; a k-DNF is one where the number of literals in one clause is at most k. It has been proved that circuit lower bounds are hard to prove using resolution and resolution with k-DNFS, and it is conjectured that they are also hard for stronger proof systems such as Frege systems~\cite{raz}.

Due to this difficulty in theoretical progress, we would like to investigate the problem empirically. SAT solvers provide a useful heuristic, as they allow us to examine small examples of the problem and find concrete rather than asymptotic bounds; due to improvements in SAT solver performance over the past years, this approach has now been put into practice. For example, Kamath et al~\cite{kamath} propose a reduction from logical design synthesis to SAT which fixes the DNF structure of the circuit and number of disjunctions used. Experiments using this reduction to find minimal circuits for 2-4 bit adders and multipliers are reported in~\cite{estrada}. Kojevnikov et al~\cite{kulikov} demonstrate a more general reduction allowing any circuit with a fixed number of gates, proving linear bounds for \(MOD^n_3\) circuits (computing whether the number of bits is equal to n modul0 3) over different Boolean bases. This was done by checking for circuit `building blocks' with 5 to 11 gates. In~\cite{kulikovlocal} Kulikov et al. used the same reduction to improve large circuit building blocks by searching for smaller versions of their sub-circuits. Specifically, linear upper bounds were proved or verified for \(SUM\), \(MOD^n_4\), and \(MOD^n_3\) by checking for circuits with up to 13 gates (they remarked that checking for existence of a Boolean circuit of size 13 was out of reach). Knuth also used SAT solvers to investigate lower bounds for \(MOD^n_3\)~\cite{knuth15}.

@@

We investigate feasible ways to find small circuits using SAT reductions. Previous research in this area has focused on a small number of target Boolean functions and a single reduction at a time. Instead, we use multiple combinations of reductions and SAT solvers to see which leads to the best performance. We test them on a range of symmetric functions whose lower bounds are already somewhat understood, in order to analyse the performance in relation to the bound.

We also search for a general relationship between solver performance and minimum circuit size by applying our methods to a sample of randomly generated functions. We then use solvers to prove a result, that a lower bound conjecture about MOD-3 functions due to Knuth~\cite{knuth15} does not hold for the de Morgan basis $U_2$.

@@

SAT solvers have previously seen success in proving mathematical theorems, most famously in 2016 when Heule et al solved the Boolean Pythagorean Triples problem~\cite{heule} and in 2020 when Brakensiek, Heule et al solved Keller's conjecture~\cite{heule2}. In general, SAT solvers have the advantage of being effective even when heuristics are not known for solving the original problem, as in the case of MCSP.

We used three SAT solvers, all based on conflict-driven clause learning (CDCL). This is an algorithm closely related to the Davis–Putnam–Logemann–Loveland (DPLL) algorithm, and which is also based on the resolution proof system~\cite{cdcl}~\cite{krajicek}. As mentioned earlier, this means that they cannot decide general circuit lower bounds with a less than exponential-length proof~\cite{raz}. However, this asymptotic bound still allows for reasonably-sized proofs for small inputs such as the ones investigated in this paper.

The general idea of CDCL is to choose an assignment to a variable, propagate it (simplify the problem by removing clauses that now evaluate to true), and repeat until either finding a satisfying assignment, or finding that our assignment makes the formula unsatisfiable: in which case, add a `learned clause' to the formula according to a procedure incorporating resolution, and then `backtrack' (remove recent assignments so the formula is no longer unsatisfiable - if this is impossible, the answer is UNSAT) before trying again.

@@

PicoSAT is a solver inspired by miniSAT 1.14, with a focus on improving low-level performance by using optimised data structures~\cite{picosat}. It was submitted in the 2010 SAT-Race, and was also one of the solvers used in~\cite{kulikov} and the main one in~\cite{kulikovlocal}.

MapleSAT is the most recent solver, with variants having won the SAT Competition 2016 and placed second in 2017. It was based off MiniSAT 2.2 but has a different branching heuristic, incorporating ideas from machine learning\cite{maplesat}.

Finally, in order to make full use of the Oxford ARC service and its distributed computing, the parallel solver Glucose-Syrup was used. This is a 2014 parallel version of the Glucose solver, which placed 4th in both the 2014 and 2015 SAT Competition. Glucose is a solver also based on MiniSAT 2.2, which uses a new heuristic to more accurately estimate the 'usefulness' of its learned clauses and regularly delete ones that are not as useful~\cite{glucose}. Glucose-Syrup is an unusual 'non-portfolio' parallel solver which runs multiple instances of the same solver in parallel, but rather than taking a divide and conquer approach, uses new heuristics for sharing learned clauses between the instances in order to speed them up~\cite{glucose-syrup}.

@@

These axioms allow the SAT solver to use `Extended Resolution' - resolving on a formula incorporating multiple literals, rather than on a single literal. This is possible since $q \leftrightarrow l_1 \lr l_2$ and $l_1$, $l_2$ may also be extension variables, allowing for larger formulae to be built~\cite{krajicek}.

@@

We investigated the relationship between the size of a circuit and the time taken for a SAT solver to confirm or deny its existence. We encoded a range of simple symmetric functions using the Kojevnikov encoding:

@@

We chose to investigate the existence of circuits up to 9 gates, as these were reliably solved within less than 9 hours on the single-processor hardware described.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.75\textwidth]{images/times/8and_kulikov.png}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing AND on 8 bits, using the Kojevnikov reduction\label{fig:8and_kulikov}}
  \includegraphics[width=0.75\textwidth]{images/times/4parity_naive.png}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing parity on 4 bits, using the Naive reduction\label{fig:4parity_naive}}
\end{figure}

@@

Different functions took different amounts of time overall, varying between seconds and hours, something also seen in Section \ref{randomresults}, and as discussed there, the time appeared to be greater when the function had a greater optimal circuit size. Performance for each function was not especially consistent between the solvers; usually they found the same things difficult, but the length of time each solver took for the difficult problems varied, shown best in Figure \ref{fig:8and_kulikov}. The extent of this can be seen in Section \ref{satresults}.

@@

@@

We took the total of the times each combination took to solve the problems from Section \ref{funcresults} and compared them as shown in Figure \ref{fig:totals}.

@@

It can be seen that MapleSAT on the Kojevnikov encoding performs the best. Apart from this, the effect of the reduction and the solver seem to be independent.

MapleSAT consistently performs better than MiniSAT and PicoSAT (probably because it is the most recent and builds upon the research that went into the other two), and PicoSAT occasionally `spikes' far worse on difficult problems than the other two, suggesting something about its modifications makes it especially unsuited for solving these reductions.

The Naive and Kojevnikov reductions both performed better than the Razborov reduction, but which was best depended on the solver used. It is possible that the efficient encoding of the Razborov reduction is detrimental to the performance of the solver. However, these results are only applicable to small values of n.

@@

We investigated the effect of adding 50,000, 100,000, and 200,000 extension variables to the encodings for existence of 9-gate circuits for \(MOD_3\) and 4-bit parity functions. The literals were chosen uniformly at random using a pseudorandom number generator simulating a uniform distribution (specifically, xorshift128+ due to the TypeScript implementation). Taking note of the results from the previous section, we used the Kojevnikov reduction and MapleSAT.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.75\textwidth]{images/extvars/boxplot_4parity_medians.png}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for 4-bit parity  \label{fig:extparity}}

  \centering
  \includegraphics[width=0.75\textwidth]{images/extvars/boxplot_4mod3_medians.png}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for \(MOD_3\)  \label{fig:extmod3}}
\end{figure}

@@

64 truth tables were generated for 4-bit inputs using the same pseudorandom generator as above. The existence of circuits of size 1 to 11 for these tables were encoded with the Kojevnikov reduction and solved using MapleSAT. The times taken to check for circuits are shown plotted against their optimal circuit size in \ref{fig:4rand_scatter} (note that 2 tables could not be fully solved within 9 hours; these functions have been omitted leading to a sample size of 62). The distribution of optimal circuit sizes is shown in \ref{fig:4rand_sizes}.

@@

While we expect a large proportion of circuits to be hard (i.e. have a large optimal circuit size) for large input sizes ~\cite{arora}, the usual counting arguments break down for such small values of n. However, using the SAT solver we can check the number of hard circuits directly. While the small sample size can account for the irregular shape of the graph in \ref{fig:4rand_sizes}, there does appear to be a peak around 10 gates. This contradicts the intuition that there should be exponentially more harder functions, and suggests that there is quite a low upper limit on the complexity of a function which only can only vary over 4 bits of input. However, this is all under the assumption that the pseudorandom generator used did output a truly random sample; in reality, it may have sampled disproportionately many easy (or hard) functions.

\subsection{Knuth's MOD-3 conjecture}

Define $MOD^{m,r}_n(x_1 \ldots x_n) = [x_1 + \ldots + x_n \equiv r $mod$ m]$, where $[b]$ is the Iverson bracket, i.e. evaluates to 1 if $b$ evaluates to true, else false.

Based on the minimum circuit sizes he found for small sizes of n, Knuth conjectured that for all $r$, and $n \geq 3$, and over a full circuit basis

\[
  minsize(MOD^{m,r}_n) = 3n - 5 - [(n+r) \equiv 0 mod 3]  
\]

He was able to confirm this for $3 \leq n \leq 5$ and all $r$, and for the case $n=6, r=0$, but not for the case $n=6, r=1$.

We decided to try and check the hypothesis for $n=6, r=0$ and $n=6, r=1$ for the De Morgan basis. This involved checking for circuits for the function up to size 11 and 12, and 12 and 13 respectively. The Glucose-Syrup solver was used, running on the Oxford ARC service with a maximum of 32 threads (so, 32 parallel instances of the server working together). In order to ensrue that the number of threads would lead to a gain in performance, rather than a drop due to the extra time needed to coordinate them, the SAT solver was timed on a smaller problem with 4, 8, 16 and 32 threads; it displayed gains in performance up to 32.

@@

For $n=6, r=0$, confirming that there is no circuit size 11 took 175.37 hours, while checking for size 12 did not terminate within the alloted 2 weeks of runtime.

@@

Realising that more precise measurements were needed for the experiments, I investigated better methods to record the runtime of programs on the command line, beginning with using the `time' command but progressing to storing the local start and end time in shell variables. I also gained a greater proficiency in using stream editors and gnuplot on the command line in order to process the data from my experiments, and learned \LaTeX  \ in order to present the results and theory necessary for my report.

@@

We have confirmed that when conducting empirical investigation of circuit lower bounds, both the SAT solver and the SAT reduction used are critical for getting results without using too much time and resources. We demonstrated some different uses of the SAT solver to investigate theoretical ideas, including how high-powered computing can be incorporated to check for bigger circuits.

@@

Building on the experiments so far, it seems feasible to search for the circuit size for all 4-bit functions (there are 65536, but many of these are degenerate functions that ignore part of their inputs, or functions that compute the same bits in a different order) and for some other sizes in order to understand the distribution of minimum sizes for small inputs. Circuits with restricted depth, different bases and higher fan-in could also be investigated, for example to verify some of the stronger lower bounds that have been proved for them.

The fact that extension axioms leading to improvements were found, even with no strategy or constraints on the axioms added, is promising for research in the topic. Even without theoretical research into strategies for adding the extension axioms, it would be interesting to do data analysis on a larger sample of randomly generated extension variables - for example, measuring the improvement in performance against the depth of circuits added or whether literals are chosen from the variables determining gate types or graph edges.

It would also be a good idea to compare more recent SAT solvers on the circuit bounding problem, analyse the reasons behind the better performance of certain ones using prior research into SAT solvers, and explore they might be customised to suit the problem better. Similarly, it would be interesting to compare the Kojevnikov reduction to even more types of reductions and analyse why SAT solvers perform better on some than others.

@@

>>>>

If functions with circuits of a given complexity could be identified by an efficient algorithm, this could be used to construct efficient learning algorithms for circuits of a smaller complexity; this would resolve one of the biggest questions in learning theory~\cite{carmosino}. Furthermore, such an algorithm could be used to break pseudorandom generators (i.e. distinguish random from pseudorandom inputs)~\cite{razborov94}. As a result, this problem has strong connections in both learning theory and cryptography. Proving circuit lower bounds would also imply derandomisation - that all randomised algorithms have a deterministic counterpart which is only polynomially slower~\cite{arora}.

A heuristic approach is to find minimal circuits for small instances of the functions. This can be used to prove new bounds~\cite{kulikovsurvey} and may lead to theoretical insights on the structure of their optimal circuits generally~\cite{williams}. Doing this for small functions computing natural functions like parity, modulo and addition allows us to gain a greater understanding of their difficulty and compute them more efficiently when using low-level code or hardware.

Finding efficient circuits is also necessary in practice when designing electronic systems. Electronic circuit design is an extensive field of its own, involving circuit simulation software to test designs, or hardware specification languages such as Verilog and circuit synthesis software to convert specifications into circuits. However, electronic circuit design is concerned with a wider range of problems (for example, analogue and time-dependent functions) and with concerns about correct physical implementation; we focus on abstract circuits.

@@

Circuits are a popular model of computation; they are non-uniform, meaning that in a family of circuits where each takes a different size of input, each circuit is independent and not constructed by a common algorithm. Notably, this means that circuits are not subject to the Space and Time Hierachy Theorems, although there is a Non-Uniform Hierarchy Theorem stating that larger circuits can compute strictly more functions than smaller ones~\cite{arora}. In the past there have been theoretical proofs of lower bounds for restricted circuit classes, such as $AC^0$ (functions computed by circuit families of $O(1)$ depth)~\cite{furst81}~\cite{ajtai83}, $AC^0[k]$ (i.e. $AC^0$ additionally with MOD-k gates)~\cite{razborov87}~\cite{smolensky87}, and monotone circuits~\cite{razborov85}. There was also a bound showing the practical impossibility of deciding bounded-size formulae in a certain second-order language~\cite{stockmeyer02}. However, no lower bounds better than $(3+1/86)n-o(n)$ have been found for general circuits over the basis $B_2$ consisting of all functions on 2 bits~\cite{find16}, while the best bound is 5n-c for the basis $U_2 = \{ \land, \lor, \neg\}$~\cite{lachish01}~\cite{iwama02}. Furthermore, there are results stating that it is impossible to prove circuit lower bounds separating P and NP by relativising methods~\cite{baker75}, algebrizing methods~\cite{aaronson09}, or (under standard cryptographic assumptions of the existence of strong one-way functions) so-called `natural proofs'~\cite{razborov94}. A formula in Disjunctive Normal Form (DNF) is one consisting of clauses (disjunctions of literals), with the final formula consisting of the conjunction of all clauses; a $k$-DNF is one where the number of literals in one clause is at most $k$. It has been proved that circuit lower bounds are hard to prove using Resolution and Resolution with $k$-DNFS, and it is conjectured that they are also hard for stronger proof systems such as Frege systems~\cite{raz}.

Due to this difficulty in theoretical progress, we would like to investigate the problem empirically. SAT solvers provide a useful heuristic, as they allow us to examine small examples of the problem and find concrete rather than asymptotic bounds; due to improvements in SAT solver performance over the past years, this approach has now been put into practice. For example, Kamath et al~\cite{kamath} propose a reduction from logical design synthesis to SAT which fixes the DNF structure of the circuit and number of disjunctions used. Experiments using this reduction to find minimal circuits for 2-4 bit adders and multipliers are reported in~\cite{estrada}.

Kojevnikov et al~\cite{kulikov} demonstrate a more general reduction allowing any circuit with a fixed number of gates, using it to prove both upper and lower bounds for both specific functions and the general case. In particular, they prove linear bounds for \(MOD^n_3\) circuits (computing whether the number of bits is equal to n modulo 3) over different Boolean bases. This was done by checking for circuit `building blocks' with 5 to 11 gates. This was the first large-scale, direct attempt to use SAT solvers to search for circuit bounds.

In~\cite{kulikovlocal} Kulikov et al used the same reduction to improve large circuit building blocks by searching for smaller versions of their sub-circuits. Specifically, linear upper bounds were proved or verified for the symmetric functions \(SUM\), \(MOD^n_4\), and \(MOD^n_3\) by checking for circuits with up to 13 gates (they remarked that checking for existence of a Boolean circuit of size 13 was out of reach). They also improved on the best known lower bound for the threshold function (which outputs true if more than k input bits are set to true, for some k). Their technique allowed them to search for improvements on circuits that were much larger than a SAT solver could deal with if taken as a whole.

Knuth also used SAT solvers to investigate lower bounds for \(MOD^n_3\)~\cite{knuth15} (solution to exercise 480). Define $MOD^{m,r}_n(x_1 \ldots x_n) = [x_1 + \ldots + x_n \equiv r \ $mod$ \ m]$, where $[b]$ is the Iverson bracket, i.e. evaluates to 1 if $b$ evaluates to true, else 0.

Based on the minimum circuit sizes he found for small sizes of n, Knuth conjectured that for all $r$, and $n \geq 3$,

\[
  size(MOD^{m,r}_n) = 3n - 5 - [(n+r) \equiv 0 \ mod \ 3]  
\]

where $size(f)$ is the size of a minimal circuit over the $B_2$ basis which computes $f$. He was able to confirm this for $3 \leq n \leq 5$ and all $r$, and for the case $n=6, r=0$, but not for the case $n=6, r=1$.

@@

We investigate feasible ways to find small circuits using SAT reductions. Previous research in this area has focused on a small number of target Boolean functions and a single reduction at a time. Instead, we use multiple combinations of reductions and SAT solvers to see which leads to the best performance, displayed in Section \ref{satresults}. In Section \ref{funcresults}, we tested them on a range of symmetric functions whose lower bounds are already somewhat understood, in order to analyse the performance in relation to the bound.

We also search for a general relationship between solver performance and minimum circuit size in Section \ref{randomresults} by applying our methods to a sample of randomly generated functions. In Section \ref{knuthresults} we then use solvers to prove a result, that a lower bound conjecture about MOD-3 functions due to Knuth~\cite{knuth15} does not hold for the de Morgan basis $U_2$.

@@

SAT solvers have previously seen success in proving mathematical theorems, most famously in 2016 when Heule et al solved the Boolean Pythagorean Triples problem~\cite{heule} and in 2020 when Brakensiek, Heule et al solved Keller's conjecture~\cite{heule2}. In general, SAT solvers have the advantage of being effective even when heuristics are not known for solving the original problem, as in our case of MCSP.

Modern SAT solvers are based on conflict-driven clause learning (CDCL). This is an algorithm closely related to the Davis–Putnam–Logemann–Loveland (DPLL) algorithm, and which is also based on the Resolution proof system~\cite{cdcl}~\cite{krajicek}. The general idea of DPLL is to choose an assignment to a variable, propagate it (simplify the problem by removing clauses that now evaluate to true), and repeat until either finding a satisfying assignment, or finding that our assignment makes the formula unsatisfiable: in which case `backtrack' (remove the last assignment until the formula is no longer unsatisfiable - if this is impossible, the answer is UNSAT) before trying again with a different assignment. This is essentially a form of depth-first tree exploration. The major difference in CDCL is that it adds a `learned clause' to the formula when it reaches an unsatisfiable assignment, according to a procedure incorporating Resolution; it backtracks until the learned clause is also satisfiable, allowing it to avoid some other unsatisfiable assignments~\cite{bayardo97}.

For this project we used 4 different SAT solvers, all based on CDCL. As mentioned earlier, this means that they cannot decide general circuit lower bounds with a less than exponential-length proof~\cite{raz}. However, this is an asymptotic bound and still allows for reasonably-sized proofs for small inputs such as the ones investigated in this paper.

@@

PicoSAT is a solver inspired by miniSAT 1.14, with a focus on improving low-level performance by using optimised data structures~\cite{picosat}. It was submitted in the 2010 SAT-Race, and was also one of the solvers used in~\cite{kulikov} and the main one in~\cite{kulikovlocal}, which is why it was chosen.

MapleSAT is the most recent solver, chosen for its good track record, with variants having won the SAT Competition 2016 and placed second in 2017. It was based off MiniSAT 2.2 but has a different branching heuristic, incorporating ideas from machine learning\cite{maplesat}.

Finally, in order to make full use of the Oxford ARC service and its distributed computing, the parallel solver Glucose-Syrup was used. This is a 2014 parallel version of the Glucose solver which placed 4th in both the 2014 and 2015 SAT Competition. Glucose is a solver also based on MiniSAT 2.2, which uses a new heuristic to more accurately estimate the `usefulness' of its learned clauses and regularly delete ones that are not as useful~\cite{glucose}. Glucose-Syrup is an unusual `non-portfolio' parallel solver which runs multiple instances of the same solver in parallel, but rather than taking a divide and conquer approach, uses new heuristics for sharing learned clauses between the instances in order to speed them up~\cite{glucose-syrup}.

These open-source SAT solvers take input in the form of text files in the DIMACS format~\cite{dimacs} and need to be run from the command line; for this project, we wrote a TypeScript library to store CNF formulae together with comments and output them correctly both in human-readable and DIMACS formats.

@@

These axioms allow the SAT solver to use `Extended Resolution' - resolving on a formula incorporating multiple literals, rather than on a single literal. This is possible since $q \leftrightarrow l_1 \lr l_2$ and $l_1$, $l_2$ may also be extension variables, allowing for larger formulae to be built~\cite{krajicek}. Some statements that require an exponential length proof in Resolution require only a polynomially long proof in Extended Resolution, e.g. a formula stating the pigeonhole principle~\cite{cook79}.

@@

We investigated the relationship between the size of a circuit and the time taken for a SAT solver to confirm or deny its existence. We used the MiniSAT, PicoSAT and MapleSAT solvers and encoded a range of simple symmetric functions using the Kojevnikov encoding:

@@

We chose to investigate the existence of circuits up to 9 gates, as during prior tests these were reliably solved using MiniSAT within less than 9 hours on the single-processor hardware described.

@@

@@

Different functions took different amounts of time overall, varying between seconds and hours, something also seen in Section \ref{randomresults}, and as discussed there, the time appeared to be greater when the function had a greater optimal circuit size. Performance for each function was not especially consistent between the solvers; usually they found the same things difficult, but the length of time each solver took for the difficult problems varied, shown clearly in Figure \ref{fig:8and_kulikov}. The extent of this can be seen in Section \ref{satresults}.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{images/times/8and_kulikov.png}
  \caption{Times taken by SAT solvers to check for existence of a circuit computing AND on 8 bits, using the Kojevnikov reduction\label{fig:8and_kulikov}}
\end{figure}

\begin{figure}[h!]
  \centering
\includegraphics[width=0.8\textwidth]{images/times/4parity_naive.png}
\caption{Times taken by SAT solvers to check for existence of a circuit computing parity on 4 bits, using the Naive reduction\label{fig:4parity_naive}}
\end{figure}

@@

We took the total of the times each combination took to solve the problems from Section \ref{funcresults} and compared them as shown in Figure \ref{fig:totals}. It can be seen that MapleSAT on the Kojevnikov encoding performs the best. Apart from this, the effect of the reduction and the solver seem to be independent.

@@

MapleSAT consistently performs better than MiniSAT and PicoSAT (probably because it is the most recent and builds upon the research that went into the other two), and PicoSAT sometimes `spikes' far more badly on difficult problems than the other two, suggesting something about its modifications makes it especially unsuited for solving these reductions.

The Naive and Kojevnikov reductions both performed better than the Razborov reduction, but which was best depended on the solver used. It is possible that the efficient encoding of the Razborov reduction is detrimental to the performance of the solver. However, these results are only applicable to small values of n; possibly at large sizes the efficient encoding would lead to a noticeably smaller search space and make the problem easier to solve.

\subsection{Extension axioms}\label{extresults}

We investigated the effect of adding 50,000, 100,000, and 200,000 extension variables to the encodings for existence of 9-gate circuits for 4-bit \(MOD_3\) and 4-bit parity functions. The literals were chosen uniformly at random using a pseudorandom number generator simulating a uniform distribution (specifically, xorshift128+ due to the TypeScript implementation). Taking note of the results from the previous section, we used the Kojevnikov reduction and MapleSAT.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{images/extvars/boxplot_4parity_medians.png}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for 4-bit parity  \label{fig:extparity}}
\end{figure}

@@

  \begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{images/extvars/boxplot_4mod3_medians.png}
  \caption{Comparison of times taken to check existence of a 9-gate circuit for \(MOD_3\)  \label{fig:extmod3}}
\end{figure}

@@

After seeing the variation in times between functions in the previous sections, we decided to investigate the overall distribution of times for functions with the same input size. We decided to investigate 4-bit functions, as Knuth did in ~\cite{knuth15} (Section 7.1.2, Table 1) but for the De Morgan basis rather than the full basis. However, while he checked all 65536 functions, we chose to take a random sample of 64 functions rather than checking all of them. This has the advantage that the experiment could be completed in a small amount of time even for larger input sizes, where checking all functions would be infeasible due to the difficulty of the problems and the large number of functions.

64 truth tables were generated for 4-bit inputs, uniformly setting the output bit to 0 or 1 for each input, using the same pseudorandom generator as above. The existence of circuits of size 1 to 11 for these tables were encoded with the Kojevnikov reduction and solved using MapleSAT.

The times taken to check for circuits are shown plotted against their optimal circuit size in \ref{fig:4rand_scatter} (note that 2 tables could not be fully solved within 9 hours; these functions have been omitted leading to a sample size of 62). The distribution of optimal circuit sizes is shown in \ref{fig:4rand_sizes}.

@@

While we expect a large proportion of circuits to be hard (i.e. have a large optimal circuit size) for large input sizes ~\cite{arora}, the usual counting arguments break down for such small values of n due to large coefficients for the circuit encoding size. However, using the SAT solver we can check the number of hard circuits directly. While the small sample size can account for the irregular shape of the graph in \ref{fig:4rand_sizes}, there does appear to be a peak around 10 gates. This contradicts the intuition that there should be exponentially more harder functions, and suggests that there is quite a low upper limit on the complexity of a function which only can only vary over 4 bits of input.

While this is all under the assumption that the pseudorandom generator used did output a truly random sample (in reality, it may have sampled disproportionately many easy or hard functions), it is encouraging that Knuth's results for all functions in~\cite{knuth15} also follow a curve similar to ours - although since his basis was $B_2$, where gates can compute more complex functions, the maximum size of any circuit was only 7 and the size with the most functions was 5. This suggests that our method could still provide a good approximation of the space of all functions, even for bigger input sizes.

\begin{figure}[!h]

  \begin{minipage}{0.45\textwidth}
  \begin{centering}
  \begin{tabular}{ |c|c| }
  \hline
  Input & Output \\ 
  \hline
  0000 & 1 \\
  0001 & 1 \\
  0010 & 1 \\
  0011 & 0 \\
  0100 & 0 \\
  0101 & 1 \\
  0110 & 0 \\
  0111 & 0 \\
  1000 & 0 \\
  1001 & 1 \\
  1010 & 1 \\
  1011 & 0 \\
  1100 & 1 \\
  1101 & 0 \\
  1110 & 1 \\
  1111 & 1 \\
  \hline
  \end{tabular}
  \end{centering}
  \caption{Example truth table for function with optimal size 11 that took 87 minutes to solve all sizes}

  \

  \begin{centering}
    \begin{tabular}{ |c|c| }
    \hline
    Input & Output \\ 
    \hline
    0000 & 0 \\
    0001 & 0 \\
    0010 & 1 \\
    0011 & 0 \\
    0100 & 0 \\
    0101 & 1 \\
    0110 & 1 \\
    0111 & 0 \\
    1000 & 1 \\
    1001 & 1 \\
    1010 & 0 \\
    1011 & 1 \\
    1100 & 0 \\
    1101 & 0 \\
    1110 & 0 \\
    1111 & 0 \\
    \hline
    \end{tabular}
    \end{centering}
    \caption{Example truth table for function with optimal size 10 that took 95 minutes to solve all sizes}

\end{minipage}\hfill
\begin{minipage}{0.45\textwidth}
  \begin{centering}
    \begin{tabular}{ |c|c| }
    \hline
    Input & Output \\ 
    \hline
    0000 & 1 \\
    0001 & 1 \\
    0010 & 0 \\
    0011 & 1 \\
    0100 & 0 \\
    0101 & 1 \\
    0110 & 1 \\
    0111 & 1 \\
    1000 & 0 \\
    1001 & 0 \\
    1010 & 1 \\
    1011 & 0 \\
    1100 & 1 \\
    1101 & 1 \\
    1110 & 1 \\
    1111 & 0 \\ 
    \hline
    \end{tabular}
    \end{centering}
    \caption{Example truth table for function with optimal size 11 that took 282 minutes to solve all sizes}

    \

    \begin{centering}
      \begin{tabular}{ |c|c| }
      \hline
      Input & Output \\ 
      \hline
      0000 & 0 \\
      0001 & 1 \\
      0010 & 0 \\
      0011 & 0 \\
      0100 & 0 \\
      0101 & 0 \\
      0110 & 0 \\
      0111 & 1 \\
      1000 & 0 \\
      1001 & 0 \\
      1010 & 0 \\
      1011 & 1 \\
      1100 & 1 \\
      1101 & 1 \\
      1110 & 1 \\
      1111 & 0 \\
      \hline
      \end{tabular}
      \end{centering}
      \caption{Example truth table for function with optimal size 10 that took 31 minutes to solve all sizes}
  \end{minipage}
\end{figure}

\subsection{Knuth's MOD-3 conjecture}\label{knuthresults}

Finally, we decided to check whether the power of SAT solvers and the available computing power had progressed enough from the previously published papers to allow us to solve new problems. For this we decided to re-investigate Knuth's conjecture about the circuit size for modulo 3 functions.

Since he was able to confirm that for the $B_2$ basis $size(MOD^{m,r}_n) = 3n - 5 - [(n+r) \equiv 0 \ mod \ 3] $ for $3 \leq n \leq 5$ and all $r$, and for the case $n=6, r=0$, but not for the case $n=6, r=1$, we decided to try and check the hypothesis for $n=6, r=0$ and $n=6, r=1$ for the De Morgan basis. This involved checking for circuits for the function up to size 11 and 12, and 12 and 13 respectively.

The Glucose-Syrup solver was used, running on the Oxford ARC service with a maximum of 32 threads (so, 32 parallel instances of the solver working together). In order to ensure that the number of threads would lead to a gain in performance, rather than a drop due to the extra time needed to coordinate them, the SAT solver was timed on a smaller problem with 4, 8, 16 and 32 threads; it displayed gains in performance up to 32.

@@

For $n=6, r=1$, confirming that there is no circuit size 12 took 175.37 hours, while checking for size 13 did not terminate within the alloted 2 weeks of runtime, corroborating Kulikov et al's account in~\cite{kulikovlocal} that solvers cannot yet handle circuits of size 13 or higher.

We also attempted to check $n=6, r=1$ for the $B_2$ basis, but again the solver did not terminate within 2 weeks.

@@

Realising that more precise measurements were needed for the experiments, I investigated better methods to record the runtime of programs on the command line, beginning with using the `time' command but progressing to storing the local start and end time in shell variables for greater accuracy, and outputting the time elapsed to text files together with additional columns of data to make it easier to understand and plot. I also gained a greater proficiency in using stream editors and gnuplot on the command line in order to process the data from my experiments, and learned \LaTeX  \ in order to present the results and theory necessary for my report.

@@

We have confirmed that when conducting empirical investigation of circuit lower bounds, both the SAT solver and the SAT reduction used are critical for getting results without using too much time and resources, and the effect that each has is not always independent. This is an area that prior work in the field had neglected. We demonstrated some different uses of the SAT solver to investigate theoretical ideas, including how high-powered computing can be incorporated to check for bigger circuits.

@@

Building on the experiments so far, it seems feasible to search for the circuit size for all 4-bit functions, as Knuth did, in order to understand the distribution of minimum sizes for small inputs and see if it differs significantly in the $U_2$ basis. Our sampling method could also be used to investigate bigger inputs, where checking all functions would be infeasible due to the sheer number, such as 9-bit inputs. For these input sizes, the counting argument proof that most circuits are exponential would apply, and we would be able to verify this experimentally. Circuits with restricted depth, different bases and higher fan-in could also be investigated, for example to verify some of the stronger lower bounds that have been proved for them.

The fact that extension axioms leading to improvements were found, even with no strategy or constraints on the axioms added, is promising for research in the topic. Even without theoretical research into strategies for adding the extension axioms, it would be interesting to do data analysis on a larger sample of randomly generated extension variables - for example, measuring the improvement in performance against the depth of circuits added, or whether literals are chosen from the variables determining gate types or those determining graph edges.

It would also be a good idea to compare how more recent SAT solvers do on the MCSP, analyse the reasons behind the better performance of certain ones using prior research into SAT solvers, and explore how they might be customised to suit the problem better. Similarly, it would be interesting to compare the Kojevnikov reduction to even more types of reductions and analyse why SAT solvers perform better on some than others.

@@

\section{Appendix}

\subsection{Graphs of solver performance when checking for existence of circuits computing various symmetric functions}
